Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-06-04 main.tex
Combined the two 6 km/h stats tables.

\def\kph{\kilo\meter\per\hour}

@@

\begin{table}
  \small
  \begin{tabular}{lll}
    \toprule
    \midrule
    \(L\) & \si{\newton\meter\second} & angular impulse of the applied
    perturbation torque \\
    \(c\) & integer & order number of perturbation \\
    \(\sigma_\delta\) & standard deviation of steer angle during Phase X \\
    \(k\) & TODO & gain of balance assist control \\
    \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
    \(v\) & \si{\meter\per\second} & forward speed (6 or 10) \\
    \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
    \bottomrule
  \end{tabular}
\end{table}

We evaluate this hypothesis using a multivariate logistic regression model.

\begin{align}
  f_{ij} | p_{ij} \sim \textrm{Bern}(p_{ij})
\end{align}

\(f_{ij}\) is the binary outcome of perturbation \(j\) of participant \(i\)
which follows a Bernoulli distribution given the probability \(p_{ij}\) that a
fall occurs. The log-odds of the probabiliy is then a linear function of our
independent variables with \(\beta\) as the intercept and \(\alpha_k\) as the
linear coefficients.

@@

We scalle all independent variables such they they have a mean of zero and a
standard deviation of one. We use cluster-mean centering, as recommended by
\cite{Enders2007}, with the clusters being an individual subject because we are
interested in the association between the state of the balance-assist system
and the outcome of the perturbation.

Cluster-mean centering shows there to be no variation between participants,
i.e. that rider skill (the only participant differentiating factor) is not a
predictor. We expected there to be a variation between participants in how well
they are able to resist the perturbation. However, this was not true. This
allowed us to use a simpler single-level logistic regression model, instead of
a multilevel model. This left use with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation. We
also include interaction effects between the balance assist state and the other
four variables.

\todo[inline]{Marten writes ``The skill variable is not included, because it
was not collected correctly for all participants.'' This needs to be addressed.
What about skill being tied to participant clusters?}

@@

The coefficient estimates for a single-level logistic regression at
6~\unit{\kilo\meter\per\hour} with gain $k=10$ are shown in Table
\ref{table:freq-coefs-6}. The angular impulse, perturbation order, and balance
assist state are all statistically significant predictors. Larger angular
impulse increases the probability to fall and enduring more perturbations or
having the balance assist on, decrease the probability to fall.  The associated
multiplicative change in odds are shown in
Table~\ref{table:freq-change-in-odds-6}.

\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates at
    6~\unit{\kilo\meter\per\hour} and gain $k=10$.}
  \label{table:freq-coefs-6}
  \begin{tabular}{lrrr}
    \toprule
    Variable                                      & Estimate          & Standard error          & p-value          \\
    \midrule
    Intercept                                     & -0.29             & 0.17                    & 0.09             \\
    Angular impulse                               & 1.69              & 0.27                    & 0.00             \\
    Perturbation order                            & -0.77             & 0.22                    & 0.00             \\
    Balance assist state                          & -0.64             & 0.27                    & 0.02             \\
    Roll angle                                    & -0.25             & 0.21                    & 0.24             \\
    Steer angle                                   & -0.14             & 0.21                    & 0.51             \\
    Balance-assist on $\times$ roll angle         & 0.52              & 0.34                    & 0.12             \\
    Balance-assist on $\times$ steer angle        & -0.41             & 0.34                    & 0.22             \\
    Balance-assist on $\times$ angular impulse    & 0.41              & 0.41                    & 0.32             \\
    Balance-assist on $\times$ perturbation order & -0.53             & 0.34                    & 0.12             \\
    \bottomrule
  \end{tabular}
\end{table}

\todo[inline]{Combine the columns of the multiplicative change in odds into the
coefficient table}

\begin{table}
  \centering
  \caption{Multiplicative change in odds of predictor variables at 6
  \unit{\kilo\meter\per\hour} and gain $k=10$ based on frequentist
         single-level logistic regression.}
  \label{table:freq-change-in-odds-6}
  \begin{tabular}{lllll}
      \textbf{Variable}                             & \textbf{Multiplicative change in odds} & \textbf{2.5\%} & \textbf{97.5\%} & \textbf{p-value} \\ \hline
      Intercept                                     & 0.75                                   & 0.53           & 1.05            & 0.09             \\
      Angular impulse                               & 5.40                                   & 3.18           & 9.16            & 0.00             \\
      Perturbation order                            & 0.46                                   & 0.30           & 0.72            & 0.00             \\
      Balance-assist on                             & 0.53                                   & 0.31           & 0.89            & 0.02             \\
      Roll angle                                    & 0.78                                   & 0.51           & 1.18            & 0.24             \\
      Steer angle                                   & 0.87                                   & 0.58           & 1.32            & 0.51             \\
      Balance-assist on $\times$ roll angle         & 1.68                                   & 0.86           & 3.29            & 0.12             \\
      Balance-assist on $\times$ steer angle        & 0.66                                   & 0.34           & 1.28            & 0.22             \\
      Balance-assist on $\times$ angular impulse    & 1.51                                   & 0.67           & 3.38            & 0.32             \\
      Balance-assist on $\times$ perturbation order & 0.59                                   & 0.30           & 1.15            & 0.12             \\
  \end{tabular}
\end{table}

@@

@@

\begin{table}
  \centering
  \caption{Multiplicative change in odds of predictor variables at 10 \unit{\kilo\meter\per\hour} based on
      frequentist single-level logistic regression.} \label{table:freq-change-in-odds}
  \begin{tabular}{lllll}
    \textbf{Variable}                             & \textbf{Multiplicative change in odds} & \textbf{2.5\%} & \textbf{97.5\%} & \textbf{p-value} \\ \hline
    Intercept                                     & 0.78                                   & 0.57           & 1.07            & 0.12             \\
    Angular impulse                               & 10.92                                  & 6.23           & 19.13           & 0.00             \\
    Perturbation order                            & 0.31                                   & 0.21           & 0.48            & 0.00             \\
    Balance-assist on                             & 0.64                                   & 0.40           & 1.03            & 0.07             \\
    Roll angle                                    & 1.31                                   & 0.85           & 2.04            & 0.22             \\
    Steer angle                                   & 0.69                                   & 0.43           & 1.10            & 0.12             \\
    Balance-assist on $\times$ roll angle         & 0.54                                   & 0.28           & 1.07            & 0.08             \\
    Balance-assist on $\times$ steer angle        & 1.76                                   & 0.88           & 3.50            & 0.11             \\
    Balance-assist on $\times$ angular impulse    & 1.59                                   & 0.68           & 3.74            & 0.29             \\
    Balance-assist on $\times$ perturbation order & 0.69                                   & 0.37           & 1.30            & 0.25             \\
  \end{tabular}
\end{table}

@@

Turning the balance-assist system on significantly \((p<0.05)\) reduces the
odds that a fall occurs while cycling at a speed of 6~\si{\kph} but at
10~\si{\kph} that is not the case \((p=0.07)\). If all other factors remain the
same at 6~\si{\kph}, the balance-assist system halves the odds that a
perturbation results in a fall. How much this changes the probability that a
fall occurs depends on the values of the other predictor variables: if the odds
that a perturbation results in a fall are a 1000:1, turning on the
balance-assist system reduces the odds to 500:1. However, the probability that
a fall occurs is only reduced from $0.999$ to $0.998$. On the other hand, if
the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability. For example, if the odds that a fall occurs
is two, halving it to one reduces the probability from $0.66$ to $0.50$. Figure
\ref{fig:probability} shows how the balance-assist system changes the fall
probability for different magnitudes of angular impulse.

\begin{figure}
  \centering
  \subcaptionbox{6~\si{\kph}}{
  \includegraphics[width=70mm]{example-image-a}
    %\includegraphics[width=\textwidth]{figures/predicted_fall_probability-6.png}
  }
  \subcaptionbox{10~\si{\kph}}{
    \includegraphics[width=70mm]{example-image-b}
  }
  \caption{Comparison of predicted fall probability for balance-assist on or
    off All other predictor variables are fixed at zero, except for the
    interaction effect between the balance-assist and the angular impulse.}
  \label{fig:probability}
\end{figure}

>>>>

\def\kph{\kilo\meter\per\hour}
\def\mps{\meter\per\second}

@@

\begin{table}
  \centering
  \caption{Statisitcal model variables.}
  \begin{tabular}{lll}
    \toprule
    Variable & Units & Description \\
    \midrule
    \(L\) & \si{\newton\meter\second} & angular impulse of perturbation torque \\
    \(c\) & integer & order number of perturbation \\
    \(k\) & TODO & gain of balance assist control \\
    \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
    \(v\) & \si{\meter\per\second} & forward speed (1.7 or 2.8) \\
    \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
    \bottomrule
  \end{tabular}
\end{table}

We evaluate this hypothesis using a multivariate logistic regression model in
the form

\begin{align}
  f_{ij} | p_{ij} \sim \textrm{Bern}(p_{ij}) \textrm{.}
\end{align}

Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) of
participant \(i\) which follows a Bernoulli distribution given the probability
\(p_{ij}\) that a fall occurs. The log-odds of the probabiliy is then a linear
function of our independent variables with \(\beta\) as the intercept and
\(\alpha_k\) as the linear coefficients.

@@

We scale all independent variables such they they have a mean of zero and a
standard deviation of one. We use cluster-mean centering, as recommended by
\cite{Enders2007}, with the clusters being an individual subject because we are
interested in the association between the state of the balance-assist system
and the outcome of the perturbation.

Cluster-mean centering shows there to be no variation between participants. We
expected there to be a variation between participants in how well they are able
to resist the perturbation. However, this was not true. This allowed us to
stick with a simple single-level logistic regression model, instead of a
multilevel model. This left use with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independen tvariables.. We also include interaction effects between the balance
assist state and the other four variables.

@@

The coefficient estimates for a single-level logistic regression at
6~\unit{\kilo\meter\per\hour} with gain $k=10$ are shown in
Table~\ref{table:freq-coefs-6}. The angular impulse, perturbation order, and
balance assist state are all statistically significant predictors. Larger
angular impulse increases the probability to fall and enduring more
perturbations or having the balance assist on, decrease the probability to
fall.  The associated multiplicative change in odds are shown in
Table~\ref{table:freq-change-in-odds-6}.

\todo{Combine tables 3 and 4 into one}

\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates at
    6~\unit{\kilo\meter\per\hour} and gain $k=10$. Multiplicative change in odds of predictor variables at 6
  \unit{\kilo\meter\per\hour} and gain $k=10$ based on frequentist
         single-level logistic regression.}
  \label{table:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & Estimate & Standard error & p-value & MCIO & 2.5\% & 97.5\% \\
    \midrule
    Intercept & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse & 1.69 & 0.27 & 0.00 & 5.40 & 3.18 & 9.16 \\
    Perturbation order & -0.77 & 0.22 & 0.00 & 0.46 & 0.30 & 0.72 \\
    Balance assist state & -0.64 & 0.27 & 0.02 & 0.53 & 0.31 & 0.89 \\
    Roll angle & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance-assist on $\times$ roll angle & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance-assist on $\times$ steer angle & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance-assist on $\times$ angular impulse & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance-assist on $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\todo{Combine tables 4 and 5 into one}

@@

\begin{table}
  \centering
  \caption{Multiplicative change in odds of predictor variables at 10
    \unit{\kilo\meter\per\hour} based on frequentist single-level logistic
    regression.}
  \label{table:freq-change-in-odds}
  \begin{tabular}{lllll}
    \toprule
    Variable & Multiplicative change in odds & 2.5\% & 97.5\% & p-value \\ \hline
    \midrule
    Intercept & 0.78 & 0.57 & 1.07 & 0.12 \\
    Angular impulse & 10.92 & 6.23 & 19.13 & 0.00 \\
    Perturbation order & 0.31 & 0.21 & 0.48 & 0.00 \\
    Balance-assist on & 0.64 & 0.40 & 1.03 & 0.07 \\
    Roll angle & 1.31 & 0.85 & 2.04 & 0.22 \\
    Steer angle & 0.69 & 0.43 & 1.10 & 0.12 \\
    Balance-assist on $\times$ roll angle & 0.54 & 0.28 & 1.07 & 0.08 \\
    Balance-assist on $\times$ steer angle & 1.76 & 0.88 & 3.50 & 0.11 \\
    Balance-assist on $\times$ angular impulse & 1.59 & 0.68 & 3.74 & 0.29 \\
    Balance-assist on $\times$ perturbation order & 0.69 & 0.37 & 1.30 & 0.25 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

Turning the balance-assist system on significantly \((p<0.05)\) reduces the
odds that a fall occurs while cycling at a speed of 1.7~\si{\mps} but at
2.8~\si{\mps} that is not the case \((p=0.07)\). If all other factors remain
the same at 1.7~\si{\mps}, the balance-assist system halves the odds that a
perturbation results in a fall. How much this changes the probability that a
fall occurs depends on the values of the other predictor variables: if the odds
that a perturbation results in a fall are a 1000:1, turning on the
balance-assist system reduces the odds to 500:1. However, the probability that
a fall occurs is only reduced from $0.999$ to $0.998$. On the other hand, if
the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability. For example, if the odds that a fall occurs
is two, halving it to one reduces the probability from $0.66$ to $0.50$. Figure
\ref{fig:probability} shows how the balance-assist system changes the fall
probability for different magnitudes of angular impulse.

\begin{figure}
  \centering
  \subcaptionbox{1.7~\si{\mps}}{
    \includegraphics[width=70mm]{example-image-a}
  }
  \subcaptionbox{2.8~\si{\mps}}{
    \includegraphics[width=70mm]{example-image-b}
  }
  \caption{Comparison of predicted fall probability for balance-assist on or
    off All other predictor variables are fixed at zero, except for the
    interaction effect between the balance-assist and the angular impulse.}
  \label{fig:probability}
\end{figure}
